---
title: "Final project"
author: "Pooja Singh"
date: "2024-04-27"
output: pdf_document
---


Loading and checking the dataset
```{r}
# load the dataset 

suppressPackageStartupMessages({
  pacman::p_load(MASS, ggplot2, Stat2Data,fastDummies,tidyverse)
})
data(MetroHealth83)
head(MetroHealth83) 
# Create a mapping of city names to numerical values
city_mapping <- setNames(1:length(unique(MetroHealth83$City)), unique(MetroHealth83$City))

# Replace 'City' with numerical values based on the mapping
MetroHealth83$City_Numeric <- as.integer(factor(MetroHealth83$City, levels = names(city_mapping)))
MetroHealth83 <- MetroHealth83 %>% select(-City)
# Verify the mapping


head(MetroHealth83)
dim(MetroHealth83)


```
Correlation matrix heat map

```{r}
# Select numeric columns for correlation analysis
numeric_cols <- MetroHealth83 %>% select_if(is.numeric)  # Select only numeric columns

# Compute the correlation matrix
cor_matrix <- cor(numeric_cols, use = "complete.obs")  # Correlation matrix
# Increase plot size to reduce overlapping labels
options(repr.plot.width = 20, repr.plot.height = 20)  # Expand plot dimensions

# Create a correlation heatmap with adjusted text and label sizes
ggcorrplot::ggcorrplot(
  cor_matrix,
  method = "square",  # Use circles to denote correlation
  lab = TRUE,  # Display correlation values
  lab_size = 2,  # Increase label size for visibility
  colors = c("blue", "white", "red"),  # Color scale for heatmap
  outline.color = "black",  # Outline for circles
  title = "Correlation Matrix for MetroHealth83"  # Set plot title
) +
  theme(
    plot.title = element_text(size = 18, hjust = 0.5),  # Adjust title size
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20),  # Increase plot margins
    axis.text.x = element_text(size = 8, angle = 45),  # Rotate x-axis labels
    axis.text.y = element_text(size = 8)  # Adjust y-axis label size
  )
```
  
  
  Full Linear regtetssion model
```{r}
# Create a linear regression model
lm_full <- lm(NumRetired ~ ., data = MetroHealth83)

# Summary of the model
summary(lm_full)  # Provides coefficients, R-squared, and p-values
```

Checking for variance inflation factor 

```{r}
# Load the necessary library for VIF
library(car)  # 'car' package has the vif() function

# Calculate the VIF for the model to check for multicollinearity
vif_values <- car::vif(lm_full)  # Calculate VIF for each predictor
print(vif_values)  # Display VIF values
```
Plotting histogram of VIF 
```{r}
library(car) 

vif_values <- car::vif(lm_full)

vif_df <- data.frame(Variable = names(vif_values), VIF = vif_values)

library(ggplot2) 
ggplot(vif_df, aes(x = Variable, y = VIF, fill = Variable)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(VIF, 2)), vjust = -0.5, size = 3) +
  theme_minimal() +
  labs(title = "Variance Inflation Factor (VIF) for Each Predictor",
       x = "Predictor Variables",
       y = "VIF Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

Summary of reduced model by eliminating predictors with greater VIF values than 30

```{r}
lm_reducedVIF<-lm(NumRetired ~ RateMDs + RateBeds  + PctChangeMedicare+ SSBChange + SSINum + SSIRate  + City_Numeric, data=MetroHealth83)
summary(lm_reducedVIF)# choose the predictors if they have vif  less than 50
```
Checking for AIC

```{r}
# Calculate AIC for the full model
aic_full <- step(lm_full)  # Lower AIC indicates a better model
print(aic_full)
```
 Summary of reduced model got from considering lowest AIC value in both direction

```{r}
lm_reducedAIC <- lm(NumRetired ~ NumBeds + NumMedicare + MedicareRate + SSBNum + SSBRate + SSINum, data = MetroHealth83)  #  reduced model with smaller AIC
summary(lm_reducedAIC)
```

 Doing Partial F-test between above two reduced model
 
```{r}
# Conduct a partial F-test to compare reduced and full models
partial_f_test <- anova(lm_reducedAIC, lm_reducedVIF)  # Compare models with ANOVA
print(partial_f_test)
```
Applying 10 fold crossvalidation for two reduced model to choose the final model

```{r}
# Load the necessary package
suppressPackageStartupMessages({
  pacman::p_load(caret)  # 'caret' package for cross-validation
})

# Define cross-validation
train_control <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation

# Create a model using cross-validation
lm_cv_AIC <- train(NumRetired ~ NumBeds + NumMedicare + MedicareRate + SSBNum + SSBRate + SSINum, data = MetroHealth83, method = "lm", trControl = train_control)

# Output cross-validation results
print(lm_cv_AIC)

# Create a model using cross-validation
lm_cv_VIF <- train(NumRetired ~ RateMDs + RateBeds  + PctChangeMedicare+ SSBChange + SSINum + SSIRate  + City_Numeric, data=MetroHealth83, method = "lm", trControl = train_control)

# Output cross-validation results
print(lm_cv_VIF)
```
Which Model Is Better?

Lower RMSE:

The first model (with 6 predictors) has a much lower RMSE, indicating that its predictions are more accurate.

Higher R-squared:

The first model explains almost all of the variance in the response variable, suggesting a good fit.

Lower MAE:

The first model also has a lower MAE, indicating smaller absolute errors on average.

What This Indicates

First Model:

Given the lower RMSE, higher R-squared, and lower MAE, this model performs better in terms of prediction accuracy and variance explanation. It suggests a good balance of predictors without overfitting or excessive complexity.

Second Model:

The significantly higher RMSE and MAE, combined with the much lower R-squared, indicate that this model has poorer prediction accuracy and explains less variance. It might be overfitting or including irrelevant predictors.

Conclusion

The first model with 6 predictors appears to be a better choice due to its lower RMSE, higher R-squared, and lower MAE, suggesting better accuracy and fit without overfitting. The second model's performance suggests possible overfitting, complexity, or the inclusion of less relevant predictors.



Analysing correlation among final six predictors model(got from AIC) that we finalize after checking 10 fold crossvalidation for both model 

Hein please add your code here for filtering out 4 predictors/6 predictors
```{r}








```

```{r}
lm_final<-lm(NumRetired ~ NumBeds + NumMedicare + SSBNum + SSINum,data=MetroHealth83)
summary(lm_final)
```

Scatter plot for final four predictors
```{r}
#scatter plot for reduced model through AIC
# Load necessary libraries
library(tidyverse)  # Data manipulation and visualization
library(GGally)  # For creating scatterplot matrices

# Create a scatterplot matrix for hospital capacity predictors
scatterplot_matrix <- GGally::ggpairs(
  MetroHealth83 %>% select(NumRetired,NumBeds,NumMedicare,SSBNum,SSINum),  # Select relevant predictors
  upper = list(continuous = "cor"),  # Show correlations in the upper triangle
  lower = list(continuous = "points"),  # Use scatter plots in the lower triangle
  diag = list(continuous = "density"),  # Density plots on the diagonal
  title = "Scatterplot Matrix for Social Security and Retirement Predictors"  # Apply the new title
) +
  theme(
    plot.title = element_text(size = 18, hjust = 0.5),  # Adjust title size and position
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 10),  # Adjust y-axis labels
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)  # Increase plot margins
  )

print(scatterplot_matrix)
```

Histogram of final model predictors 
```{r}
# Load necessary libraries
library(tidyverse)  # Data manipulation and visualization
library(patchwork)  # For combining multiple plots

# Define the histograms


plot1 <- ggplot(MetroHealth83, aes(x = NumBeds)) +
  geom_histogram(bins = 10, fill = "orange", color = "black") +
  labs(
    title = "Distribution of no. of Hospital Beds",
    x = "No. of Hospital Beds",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),  # Adjust title size and position
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Rotate x-axis labels
    axis.title.x = element_text(size = 9),  # Adjust x-axis label size
    axis.title.y = element_text(size = 9),
    plot.margin = margin(t = 20, r = 30, b = 30, l = 30)  # Increase plot margins
  )
plot2 <- ggplot(MetroHealth83, aes(x = NumMedicare)) +
  geom_histogram(bins = 10, fill = "orange", color = "black") +
  labs(
    title = "Distribution of no. of Medicare recipients",
    x = "No of Medicare recipients",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),  # Adjust title size and position
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Rotate x-axis labels
    axis.title.x = element_text(size = 9),  # Adjust x-axis label size
    axis.title.y = element_text(size = 9),
    plot.margin = margin(t = 30, r = 30, b = 30, l = 30)  # Increase plot margins
  )

plot3 <- ggplot(MetroHealth83, aes(x = SSBNum)) +
  geom_histogram(bins = 10, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Social Security Recipients",
    x = "No. of Social Security Recipients",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),  # Adjust title size and position
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Rotate x-axis labels
    axis.title.x = element_text(size = 9),  # Adjust x-axis label size
    axis.title.y = element_text(size = 9),
    plot.margin = margin(t = 30, r = 30, b = 30, l = 30)  # Increase plot margins
  )
plot4 <- ggplot(MetroHealth83, aes(x = SSINum)) +
  geom_histogram(bins = 10, fill = "orange", color = "black") +
  labs(
    title = "Distribution of Supplemental Security Income Recipients",
    x = "No. of Supplemental Security Income Recipients",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),  # Adjust title size and position
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Rotate x-axis labels
    axis.title.x = element_text(size = 9),  # Adjust x-axis label size
    axis.title.y = element_text(size = 9),
    plot.margin = margin(t = 30, r = 30, b = 30, l = 30)  # Increase plot margins
  ) 
# Combine the plots into a single row using patchwork
combined_plot <- plot1 + plot2 + plot3 + plot4  # Combine plots
combined_plot
```
Checking for assumption of final model

```{r}
# Apply square root transformation to the response and predictors
final_model_sqrt <- lm(sqrt(NumRetired) ~ sqrt(NumBeds) + sqrt(NumMedicare) + sqrt(SSBNum) + sqrt(SSINum), data = MetroHealth83)

# Plot Residuals vs Fitted
plot(final_model_sqrt, which = 1)

# QQ plot of residuals
plot(final_model_sqrt, which = 2)

# Alternatively, you can create both plots side by side using the `par()` function for layout.
par(mfrow=c(1, 2))  # Set up the graphics layout for 1 row and 2 columns
plot(final_model_sqrt, which = 1)  # Residuals vs Fitted plot
plot(final_model_sqrt, which = 2)  # QQ plot
par(mfrow=c(1, 1))  # Reset to default layout# Apply square root transformation to the response and predictors
final_model_sqrt <- lm(sqrt(NumRetired) ~ sqrt(NumBeds) + sqrt(NumMedicare) + sqrt(SSBNum) + sqrt(SSINum), data = MetroHealth83)

# Plot Residuals vs Fitted
plot(final_model_sqrt, which = 1)

# QQ plot of residuals
plot(final_model_sqrt, which = 2)

# Alternatively, you can create both plots side by side using the `par()` function for layout.
par(mfrow=c(1, 2))  # Set up the graphics layout for 1 row and 2 columns
plot(final_model_sqrt, which = 1)  # Residuals vs Fitted plot
plot(final_model_sqrt, which = 2)  # QQ plot
par(mfrow=c(1, 1))  # Reset to default layout

```



